\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
included comments by OS and JH.
was not able to constrain the ERSPC hazard ratio so that it constantly descreases

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician is their patient's risk of an event. However, methods
relying on time matching or risk-set sampling (including Cox regression)
eliminate the baseline hazard from the estimating function. As a
consequence, the focus has been on reporting hazard ratios instead of
survival curves. Indeed, reporting patient risk requires a separate
estimation of the baseline hazard. Using case-base sampling, Hanley \&
Miettinen (2009) explained how parametric hazard functions can be
estimated in continuous-time using logistic regression. Their approach
naturally leads to estimates of the survival function that are
smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis. We
illustrate all these features through three different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

The semiparametric Cox model has become the default approach to survival
analysis even though Cox himself later suggested he would prefer to
model the hazard function directly. In a 1994 interview with Professor
Nancy Reid, Sir David Cox was asked how he would model a set of censored
survival data, to which he responded: ``I think I would normally want to
tackle problems parametrically \ldots{} and if you want to do things
like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' \citep{reid1994conversation}.
Indeed, the most relevant quantity in a clinical setting is often the 5-
or 10-year risk of experiencing a certain event given the patient's
particular circumstances. Unfortunately, the most reported metric from a
Cox model is the (potentially time-dependent) hazard ratio (HR), which
ignores the duration of follow-up and is subject to selection bias
\citep{hernan2010hazards}. The covariate-adjusted survival curve
overcomes these limitations, and it is arguably a more important summary
measure to report than the HR. While stepwise survival curves can be
computed with the Cox model, they require a second step to separately
estimate the baseline hazard \citep{breslow1972discussion}.

Several authors have since pursued fully parametric approaches that made
the fitting of smooth survival curves much more transparent and
intuitive through generalized linear models. The key feature of these
procedures is splitting the time axis into discrete intervals. Whitehead
\citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
model and a Poisson regression with a parameter for each event time;
Carstensen \citeyearpar{copenhagen2012needs} provides a nice exposition
of this equivalence with a real data example and supporting R code for
computing standard errors. Arjas \& Haara
\citeyearpar{arjas1987logistic} and Efron
\citeyearpar{efron1988logistic} treated each patient-day as a Bernoulli
random variable with probability equal to the discrete hazard rate. A
potential issue with these approaches is that the number of time bins
need to be chosen by the data analyst. On the one hand, a fine grouping
of times may result in few (or none) events per interval, which then
leads to instability in the Newton-Raphson procedure for estimation
\citep[Section 4.8]{kalbfleisch2011statistical}. On the other hand, a
coarse grouping could potentially mask nonlinear trends in the hazard
function.

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (``person-moments'') in continuous time from all observed
follow-up experience constituting the study base. By doing so, they
obtained a likelihood expression for the hazard function that is
equivalent to that of logistic regression. More specifically, all
person-moments when the event of interest occurred are selected as the
case series, complemented by a randomly sampled base series of
person-moments serving as controls. This approach allows flexible
modeling of the hazard function by including time as a covariate
(e.g.~using splines or general additive models). Furthermore,
time-dependent covariates can be modeled through interactions with time.
In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
well-understood logistic regression for directly modeling the hazard
function, without requiring a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements and extends the Hanley \&
Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
parametric hazard models and covariate-adjusted survival curves using
the familiar interface of the \code{glm} function. Our implementation
includes extensions to other models such as penalized regression for
variable selection and competing-risk analysis. In addition, we provide
functions for exploratory data analysis and visualizing the estimated
quantities such as the hazard function, survival curve, and their
standard errors. The ultimate goal of our package is to make fitting
flexible hazards accessible to more end users with the hope that they
will favor reporting absolute risks and survival curves over hazard
ratios.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing R packages that implement comparable
features as \pkg{casebase}. Next, we provide some details about the
implementation of case-base sampling in our package, and we give a brief
survey of its main functions. This is followed by three case studies
that illustrate the flexibility and capabilities of \pkg{casebase}. We
show how the same framework can be used for non-linear functions of time
and non-proportional hazards, competing risks, and variable selection
via penalized regression. Finally, we end the article with a discussion
of the results and of future directions.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person-moments}. These person-moments
are indexed by both an individual in the study and a time point, and
therefore each person-moment has a covariate profile, an exposure
status, and an outcome status (i.e.~whether the event happened) attached
to it. By estimating the probability of an event occurring at a
particular person-moment, we can extract information about hazards and
survival.

Therefore, we start by sampling all person-moments at which the event
occurred; this collection of person-moments is what Hanley \& Miettinen
call the \emph{case series}. The incidence of the case series is
dictated by the hazard function of interest. Next, we sample a finite
number of person-moments at which no event occurred; this second
collection of person-moment is what Hanley \& Miettinen call the
\emph{base series}. The sampling mechanism for the base series is left
at the discretion of the user, but in practice we find that sampling
uniformly from the study base provides both simplicity and good
performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
\(N_i(t)\) count the person-moments from the \emph{case series}.

To complement the case series, we sample person-moments for the base
series. To do so, we model the base series sampling mechanism using
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(dR_i(t) = 1\) constitute the base series. We
note that the same individual can contribute multiple person-moments to
the base series. The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then
counts both the case and base series person-moments contributed by
individual \(i\). As mentioned above, the processes \(R_i(t)\) are
typically defined by the user via its intensity function \(\rho_i(t)\).
The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we can define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\), and where \(t_i\) is the event time for individual \(i\).
However, the integral over time makes the computation and maximization
of \(L_0(\theta)\) challenging.

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual likelihood contributions of
the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating function for \(\theta\) as
follows: \begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}. \label{eq:lik-function}
\end{equation} When a logarithmic link function is used for modeling the
hazard function, the above expression is of a logistic regression form
with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
selected in the case-base sampling mechanism are person-moments, rather
than individuals, and the parameters to be estimated are hazards or
hazard ratios rather than odds or odds ratios. Generally, an individual
can contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, their sampling strategy corresponds
essentially to a time-homogeneous Poisson process with hazard equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
in the base series, and \(B\) is the total population-time for the study
base (e.g.~the sum of all individual follow-up times). More complex
examples are also possible; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the intensity functions for the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. Non-uniform sampling mechanisms can
increase the efficiency of the case-base estimators.

@@

Let \(g(t; X)\) be the linear predictor such that
\(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\):

@@

\hypertarget{competing-risk-analysis}{%
\subsection{Competing-risk analysis}\label{competing-risk-analysis}}

Case-base sampling can also be used in the context of competing-risk
analysis. Assuming there are \(J\) competing events, we can show that
each person-moment's contribution to the likelihood is of the form

\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
hazard function. As may be expected, this functional form is similar to
the terms appearing in the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

To perform variable selection on the regression parameters
\(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
to the likelihood and optimise the following equation: \begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation} where \(\ell\left(\theta\right) = \log L(\theta)\) is the
log of the likelihood function given in \eqref{eq:lik-function},
\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled by the
non-negative regularization parameters \(\lambda\) and \(\alpha\), and
\(w_j\) is the penalty factor for the \(j\)th covariate. These penalty
factors serve as a way of allowing parameters to be penalized
differently. For example, we could set the penalty factor for time to be
0 to ensure it is always included in the selected model.

@@

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
provides functionality for competing-risk analysis and multistate
modelling. The package \CRANpkg{cmprsk} provides methods for
cause-specific subdistribution hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIs.

Finally, several packages include functions to estimate the survival
function and the CI. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explicitly modeled. Using this estimate, the survival function and
the CI can be obtained using their functional relationships (see
Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
\pkg{survival}. Our package \pkg{casebase} also follows this approach
for both single-event and competing-risk analyses. The second category
outlined above broadly corresponds to semi-parametric models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
functions). Packages that implement this approach include
\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the survival function, whereas the second category often produces
estimates in the form of stepwise functions.

@@

To illustrate these functions, we will use data from the European
Randomized Study of Prostate Cancer Screening (ERSPC)
\citep{schroder2009screening} which was extracted using the approach
described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
dataset is available through the \pkg{casebase} package. It contains the
individual observations for 159,893 men from seven European countries,
who were between the ages of 55 and 69 years when recruited for the
trial.

@@

\[ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
From this, we can derive the event-specific subdensities:

@@

\begin{Schunk}
\begin{Sinput}
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 100)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
anova(fit, test = "LRT")
\end{Sinput}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                                        Df Deviance Resid. Df Resid. Dev
#> NULL                                                   54539       6059
#> pspline(Follow.Up.Time, df = 2)         7    269.1     54532       5790
#> ScrArm                                  1      8.0     54531       5782
#> pspline(Follow.Up.Time, df = 2):ScrArm  7      9.1     54524       5773
#>                                        Pr(>Chi)    
#> NULL                                               
#> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
#> ScrArm                                   0.0046 ** 
#> pspline(Follow.Up.Time, df = 2):ScrArm   0.2454    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

@@

\[
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
\] where \(\lambda(t|\cdot;\hat{\beta})\) is the hazard rate as a
function of the variable \(t\) (which is usually time, but can be any
other continuous variable), \(x_1\) is the value of \(X\) for the
exposed group, \(x_0\) is the value of \(X\) for the unexposed group,
\(\mathbf{Z}\) are other covariates in the model which are equal to
\(\mathbf{z_1}\) in the exposed and \(\mathbf{z_0}\) in the unexposed
group, and \(\hat{\beta}\) are the estimated regression coefficients. As
indicated by the formula above, it is most instructive to plot the
hazard ratio as a function of a variable \(t\) only if there is an
interaction between \(t\) and \(X\). Otherwise, the resulting plot will
simply be a horizontal line across time.

@@

In this case study, we show how case-base sampling can be used in the
context of a competing-risk analysis. For illustrative purposes, we use
the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is now available as part of the
\pkg{casebase} package.

>>>>

\abstract{%
In clinical studies of time-to-event data, a quantity of interest to the
clinician is their patient's risk of an event. However, methods relying
on time matching or risk-set sampling (including Cox regression)
eliminate the baseline hazard from the estimating function. As a
consequence, the focus has been on reporting hazard ratios instead of
survival or cumulative incidence curves. Indeed, reporting patient risk
or cumulative incidence requires a separate estimation of the baseline
hazard. Using case-base sampling, Hanley \& Miettinen (2009) explained
how parametric hazard functions can be estimated in continuous-time
using logistic regression. Their approach naturally leads to estimates
of the survival or risk function that are smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can also be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis. We
illustrate all these features through three different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

The semiparametric Cox model has become the default approach to survival
analysis even though Cox himself later suggested he would prefer to
model the hazard function directly. In a 1994 interview with Professor
Nancy Reid, Sir David Cox was asked how he would model a set of censored
survival data, to which he responded: ``I think I would normally want to
tackle problems parametrically \ldots{} and if you want to do things
like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' \citep{reid1994conversation}.
Indeed, the most relevant quantity in a clinical setting is often the 5-
or 10-year risk of experiencing a certain event given the patient's
particular profile. However, the most reported metric from a Cox model
is the (potentially time-dependent) hazard ratio (HR). The
covariate-conditional survival curve is arguably a more important
summary measure to report than the HR. While stepwise survival curves
can be computed with the Cox model, they require a second step to
separately estimate the baseline hazard \citep{breslow1972discussion}.

Several authors have since pursued fully parametric approaches that made
the fitting of smooth survival curves more transparent and intuitive
through generalized linear models. The key feature of these procedures
is splitting the time axis into discrete intervals. Whitehead
\citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
model and a Poisson regression with a parameter for each event time;
Carstensen \citeyearpar{copenhagen2012needs} provides an exposition of
this equivalence with a real data example and supporting R code for
computing standard errors. Arjas \& Haara
\citeyearpar{arjas1987logistic} and Efron
\citeyearpar{efron1988logistic} treated each patient-day as a Bernoulli
random variable with probability equal to the discrete hazard rate. A
potential issue with these approaches is that the number of time bins
need to be chosen by the data analyst. On the one hand, a fine grouping
of times may result in few (or none) events per interval, which then
leads to instability in the Newton-Raphson procedure for estimation
\citep[Section 4.8]{kalbfleisch2011statistical}, and large long-format
datasets. On the other hand, a coarse grouping could potentially mask
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (``person-moments'') in continuous time from all observed
follow-up experience constituting the study base. By doing so, they
obtained a likelihood expression for the hazard function that is
equivalent to that of logistic regression with an offset term. More
specifically, all person-moments when the event of interest occurred are
selected as the case series, complemented by a randomly sampled base
series of person-moments serving as controls. This approach allows
flexible modeling of the hazard function by including functions of time
as covariates (e.g.~using splines or generalized additive models).
Furthermore, time-dependent covariates can be modeled through
interactions with time. In short, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} use the well-understood logistic
regression for directly modeling the hazard function, without requiring
a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements and extends the Hanley \&
Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
parametric hazard models and covariate-conditional survival curves using
the familiar interface of the \code{glm} function. Our implementation
includes extensions to other models such as penalized regression for
variable selection and competing risk analysis. In addition, we provide
functions for exploratory data analysis and visualizing the estimated
quantities such as the hazard function, survival curve, and their
standard errors. The ultimate goal of our package is to make fitting
flexible hazards accessible to end users who favor reporting absolute
risks and survival curves over hazard ratios.

In what follows, we first recap some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing R packages that implement comparable
features to \pkg{casebase}. Next, we provide some details about the
implementation of case-base sampling in our package, and we give a brief
survey of its main functions. This is followed by three case studies
that illustrate the flexibility and capabilities of \pkg{casebase}. We
show how the same framework can be used for non-linear functions of time
and non-proportional hazards, competing risks, and variable selection
via penalized regression. Finally, we end the article with a discussion
of the results and of future directions.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person-moments}. These person-moments
are indexed by both an individual in the study and a time point, and
therefore each person-moment has a covariate profile and an outcome
status (i.e.~whether the event happened) attached to it. By estimating
the probability of an event occurring at a particular person-moment, we
can extract information about hazards and survival.

Therefore, we start by assembling all person-moments at which the event
occurred; this collection of person-moments is what Hanley \& Miettinen
call the \emph{case series}. The incidence of the case series is
dictated by the hazard function of interest. Next, we sample a finite
number of person-moments (blinded to case moments); this second
collection of person-moments is what Hanley \& Miettinen call the
\emph{base series}. The sampling mechanism for the base series is left
at the discretion of the user, but in practice we find that sampling
uniformly from the study base provides both simplicity and good
performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
independent censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
The processes \(N_i(t)\) count the person-moments from the \emph{case
series}.

To complement the case series, we sample person-moments for the base
series. To do so, we specify the base series sampling mechanism as
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(\,\textrm dR_i(t) = 1\) constitute the base
series. We note that the same individual can contribute multiple
person-moments to the base series. The process
\(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and base
series person-moments contributed by individual \(i\). As mentioned
above, the processes \(R_i(t)\) are specified by the user via its
intensity function \(\rho_i(t)\). The process \(Q_{i}(t)\) is
characterized by
\(E[\,\textrm dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)\,\textrm dt + \rho_i(t)\,\textrm dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we can define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\), and where \(t_i\) is the event time for individual \(i\).
However, the integral over time makes the computation and maximization
of \(L_0(\theta)\) challenging using standard software.

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual partial likelihood
contributions of the form
\[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, an estimating function for \(\theta\) can be composed of
these contributions as: \begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation} When a logarithmic link function is used for modeling the
hazard function, the above expression is of a logistic regression form
with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
selected in the case-base sampling mechanism are person-moments, rather
than individuals, and the parameters to be estimated are hazards or
hazard ratios rather than odds or odds ratios. Generally, an individual
can contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the corresponding partial
likelihood score function has mean zero at the true value
\(\theta=\theta_0\), and that the resulting estimator \(\hat{\theta}\)
is asymptotically normally distributed.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, their sampling strategy corresponds to a
time-homogeneous Poisson process with intensity equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled base series
\emph{person-moments}, and \(B\) is the total population-time for the
study base (e.g.~the sum of all individual follow-up times). More
complex examples are also possible; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the intensity functions for the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. Non-uniform sampling mechanisms can
increase the efficiency of the resulting maximum partial likelihood
estimators.

@@

Here we show that certain named distributions such as exponential,
Weibull or Gompertz can be fit using our framework, though we are not
restricted to such models. Let \(g(t; X)\) be the linear predictor such
that \(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead
to different parametric hazard models. The simplest of these models is
the one-parameter exponential distribution which is obtained by taking
the hazard function to be constant over the range of \(t\):

@@

\hypertarget{competing-risk-analysis}{%
\subsection{Competing risk analysis}\label{competing-risk-analysis}}

Case-base sampling can also be used in the context of competing risk
analysis. Assuming there are \(J\) competing events, we can show that
each sampled person-moment's contribution to the partial likelihood is
of the form

\[\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
hazard function. As may be expected, this functional form is similar to
the terms appearing in the likelihood function for multinomial
regression with an offset
term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}

@@

To perform variable selection on the regression parameters
\(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
to the likelihood and optimise the following equation: \begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
\end{equation} where \(\ell\left(\theta\right) = \log L(\theta)\) is the
log of the likelihood function given in \eqref{eq:lik-function},
\(p_{\lambda,\alpha}(\theta_j)\) is a penalty term controlled by the
non-negative regularization parameters \(\lambda\) and \(\alpha\), and
\(w_j\) is the penalty factor for the \(j\)th covariate. These penalty
factors serve as a way of allowing parameters to be penalized
differently. For example, we could set the penalty factor for time to be
0 to ensure it is always included in the selected model.

@@

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
provides functionality for competing risk analysis and multistate
modelling. The package \CRANpkg{cmprsk} provides methods for
cause-specific subdistribution hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIs.

Finally, several packages include functions to estimate the survival
function and the CI. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explicitly modeled. Using this estimate, the survival function and
the CI can be obtained using their functional relationships (see
Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
\pkg{survival}. Our package \pkg{casebase} also follows this approach
for both single-event and competing risk analyses. The second category
outlined above broadly corresponds to semi-parametric models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
functions). Packages that implement this approach include
\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the survival function, whereas the second category often produces
estimates in the form of stepwise functions.

@@

To illustrate these functions, we will use data from the European
Randomized Study of Prostate Cancer Screening (ERSPC)
\citep{schroder2009screening} which was extracted using the approach
described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
dataset is available through the \pkg{casebase} package. It contains the
recreated individual observations for 159,893 men from seven European
countries, who were between the ages of 55 and 69 years when recruited
for the trial.

@@

\[ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
From this, we can derive the cause-specific subdensities:

@@

\begin{Schunk}
\begin{Sinput}
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 10)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
anova(fit, test = "LRT")
\end{Sinput}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                                        Df Deviance Resid. Df Resid. Dev
#> NULL                                                    5939       3619
#> pspline(Follow.Up.Time, df = 2)         7    246.6      5932       3373
#> ScrArm                                  1      5.6      5931       3367
#> pspline(Follow.Up.Time, df = 2):ScrArm  7      7.9      5924       3359
#>                                        Pr(>Chi)    
#> NULL                                               
#> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
#> ScrArm                                    0.018 *  
#> pspline(Follow.Up.Time, df = 2):ScrArm    0.343    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

@@

\[
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \hat{\beta})}
\] where \(\lambda(t|\cdot;\hat{\beta})\) is the hazard rate as a
function of the variable \(t\) (which is usually time, but can be any
other continuous variable), \(x_1\) is the value of \(X\) for the
exposed group, \(x_0\) is the value of \(X\) for the unexposed group,
\(\mathbf{Z}\) are other covariates in the model which are equal to
\(\mathbf{z}\) in both the exposed and unexposed group, and
\(\hat{\beta}\) are the estimated regression coefficients. As indicated
by the formula above, it is most instructive to plot the hazard ratio as
a function of a variable \(t\) only if there is an interaction between
\(t\) and \(X\). Otherwise, the resulting plot will simply be a
horizontal line across time.

@@

In this case study, we show how case-base sampling can be used in the
context of a competing risk analysis. For illustrative purposes, we use
the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is now available as part of the
\pkg{casebase} package.
